Tradition and progress. The Roman World in the Digital Age - seen through Inscriptions

### 2016-04-04 keynote/eck/Tradition and progress.tex
Eck's paper converted to LaTeX2e and added to the pdfs folder (no bibliography, though; refs are still in footnote stile -- and so they will remain, afaic)

% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US

% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex

\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

\usepackage{teubner}
% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{Tradition and progress. The Roman World in the Digital Age - seen through Inscriptions}
%\author[]{Werner Eck\corref{first}}
\author[koe]{Werner Eck\corref{first}}
\address[koe]{Universität zu Kölln} 
\cortext[first]{Corresponding author. Email: Werner.Eck@uni-koeln.de}
\begin{document}
\maketitle
%\begin{abstract}
%\end{abstract}

Epigraphy has caught up with the digital age. If this point needs to be proven, 
this very conference speaks for itself.  Its celebration in Rome does not happen 
by chance. Indeed, the first real attempt to create a collective compilation and 
presentation of all the Latin (and Greek) inscriptions in a single database or 
as a connection of the existing databases was made here in Rome in 1997 through 
Silvio Panciera during the 11\textsuperscript{th} International Congress of Greek 
and Latin Epigraphy\footnote{See S. Panciera, Informatica, in: idem, Epigrafi, epigrafia, epigrafisti, vol. 2, Rome 2006, 1913ff. – I would like to thank Aitor Blanco Perez for the translation of the paper.}. His fundamental initiative set in motion a process, which 
step by step will lead to a complete database. Very important steps in this direction 
have been made in the meantime and the development is progressing with particular 
emphasis at the Sapienza thanks to the collaboration between traditional epigraphists 
such as Silvia Orlandi and IT-people, who are affiliated to the Centro Linceo Beniamino 
Segre. These two and all the other partners in many European countries have to 
be thanked for these present steps and we wish them both courage and success in 
the following phase.

Whoever is dealing with inscriptions, at least since the 16\textsuperscript{th} 
century, has to deal with a continuously growing mass phenomenon. One of the reasons, 
which instigated Mommsen to establish the \textit{CIL}, was the amount of epigraphic 
texts, which were known at that time but had not been collected in one single place 
but were, instead, spread in countless manuscripts and publications. Even when 
one looks into a representative selection such as the \textit{Inscriptiones Latinae 
Selectae }by Hermann Dessau – completed in 1916 – there are more than 9500 
texts. Since the time of Mommsen and Dessau, however, the number of epigraphic 
documents has increased to a size previously unknown. On the website of EAGLE one 
can read of more than 1.5 million items, currently scattered across 25 EU countries 
as well as the east and south Mediterranean, which should be collected in the EAGLE 
database. In order to illustrate this growth in a more accessible way, let me refer 
to the following example. Since the conclusion of the first edition of \textit{CIL 
II}, namely with the supplement in 1892, the number of inscriptions from the Iberian 
Peninsula has tripled. For some types of inscriptions, the increase is even bigger. 
When Herbert Nesselhauf collected all the military diplomata in 1935, he could 
put together a total of 187 documents in \textit{CIL XVI. }Today, we know more 
than 1100 such texts, i.e. an increase of 7 times. Not in all regions of the early 
Roman Empire and for all types of inscriptions is the increase as dramatic as in 
the aforementioned examples. Even so, a doubling can be identified in many Roman 
places. Just to mention some examples: in Sarmizegetusa, the first Roman colony 
following the conquest of Dacia, around 330 inscriptions were collected in the 
different volumes of \textit{CIL III.} Since then, more than the same number of 
new texts has appeared, perhaps around 400\footnote{According to information by Ioan Piso}. The growth is almost always dramatic 
when intensive excavations lead to a massive increase in the amount of inscriptions 
available, as somehow happens in the cases of Sagalassos or Perge in Pamphylia\footnote{In Sagalassos almost all the new inscriptions are in Greek}. 
Perge provided only 4 Latin inscriptions for the \textit{CIL III} and today there 
are at least 42\footnote{See the Clauss-Slaby database}. Even more dramatic is the growth in Caesarea Maritima. In \textit{CIL 
}there were only 3 Latin inscriptions known from the capital of the province of 
Judea, while now we have in the Corpus Inscriptionum Iudaeae/Palaestinae more than 
270 Latin documents\footnote{W. Eck, Comprendere contesti: Il nuovo multilingue Corpus Inscriptionum Iudaeae / Palaestinae, in: Rationes Rerum 1, 2013, 17 ff.}.

Each of us is familiar with this massive increase in epigraphic monuments and we 
are reminded of this every year by the always more extensive volumes of \textit{AE} and \textit{SEG}. Even if it is impossible to provide an exact number of all the 
Latin inscriptions, one can still take as a starting point the fact that there 
are approximately 500.000 texts available for our work, including the so-called 
\textit{instrumentum domesticum}. There are more than 495.000 inscriptions in the 
Database Clauss-Slaby. Around 72.000 we can find in the Heidelberg Database and 
around the same number in the Epigraphic Database of Rome, from which the Project 
EAGLE emerged, not to mention other participants as, among others, \textit{Hispania 
Epigraphica }or the Bari Epigraphic Database \footnote{See the individual databases}. With regard to the Greek inscriptions, 
it is much more difficult to give an approximate number because their digital record 
is not as advanced as in the case of the Latin inscriptions. As for inscriptions 
in other languages – such as Punic, Hebrew-Aramaic, Demotic or Nabatean – let 
me exclude them from this discussion, even though they are just as part of the 
epigraphic tradition of the Greek and Roman antiquity as Latin and Greek inscriptions 
– we should not forget this very important fact\footnote{\label{fn:7}At the 14. International Congress of Greek and Latin Epigraphy there was a special section for this phenomenon: W. Eck, Ph. Hyse, J. Price, St. J. Seidlmayer, I.-X. Adiego, in: ÖFFENTLICHKEIT – MONUMENT – TEXT. XIV Congressus Internationalis Epigraphiae Graecae et Latinae Akten, hg. W. Eck - P. Funke, Berlin 2014, 159 ff}. Our work as historians or philologists 
is based on all those texts. But, the bigger the number of documents providing 
information is, the more difficult it becomes to find all texts. Even more difficult 
is to collect selectively those texts related to the topic on which one wants to 
work. And yet, the complete collection is crucial for the scientific result.

@@

Is that different today because there are numerous databases available? At least 
with regard to the Latin inscriptions almost all of them are collected, although 
with very different individual information. With few clicks, we can find an answer 
for many questions in few seconds – sometimes one has to wait a bit longer, since 
the server does not work properly or some other problems exist. Indeed, as an already 
deceased colleague once told me long before the digital period in 1973: \textit{Pazienza} 
is the greatest virtue of the epigraphist. Today we are ready to forget this virtue 
thanks to our computers. At any rate, most of the relevant material available in 
databases appears quickly on the screen. It is not necessary to give examples. 
And yet, something which we all know: it is by no means guaranteed that we have 
found absolutely all the relevant material which exists; the reason for these gaps 
are inter alia, that not a small part of our epigraphic testimonies survives only 
as fragments and can be restored in different ways. For example, if we would try 
to find all the \textit{quaestores }of the province of Achaea in the imperial age, 
we would not find the following fragmentary inscription from Corinth, the capital 
of the province, because the text has read as follows\footnote{\emph{CIL III} 537}:

\begin{quote}

{\it Q(uinto) Vili[o --] / Titia[no] / Quadra[to --] / IIIIvir(o) vi[ar(um) cur(andarum), trib(uno)] / la\mqs{t}icl(avio) \mqs{le}g(ionis) [--, \textbf{leg(ato) prov(inciae)] / Ach(aiae)}, ob ius[titiam et fidem,] / quam circa [universam] / Achaia[m exhibuit,] / qu\mqs{a}m pro pr(aetore) [administravit.] / \mqs{M}(arcus) An\mqs{to}nius M(arci) [f(ilius)] / Nigrinus [--].}
\end{quote}

Our search in a database would have recorded a proconsular legate of Achaea whose 
function in the inscription is nevertheless restored. The restoration of the position 
of provincial legate is not impossible, but it is also not very likely. The restoration 
of the office of quaestor:\textit{ [-- quaest(ori) prov(inciae)] / Ach(aiae) }instead 
of \textit{[-- leg(ato) prov(inciae)] / Ach(aiae) }fits in the lacuna much better\footnote{In the database Clauss-Slaby there are around 68 examples in Latin inscriptions, in the database of the Packard Institute on can find 10 more in Greek texts.}. 
For a young senator who had completed an office of the \textit{vigintiviri }and 
then a tribunate in the legion, the \textit{quaestura }should actually follow in 
a normal career and not the position of a proconsular legate under a proconsul. 
Of course, we know that exceptionally young senators already before the \textit{quaestura 
}were acting as legates of a proconsul in his province. In total, there are only 
four of such cases. By contrast, much more senators state in their inscriptions 
that they have held the office of provincial \textit{quaestor} as\textit{ pro praetore}. 
The \textit{pro pr(aetore) }preserved in the inscription could therefore refer 
to the \textit{quaestura}. As the \textit{quaestura }normally follows the\textit{ 
}viginvirate and the legionary tribunate, this rule should automatically be observed 
in a fragmentary career\footnote{See now the new reconstruction of the inscription by W. Eck, Quaestor, nicht legatus provinciae Achaiae. Zu einer Inschrift des Q. Vilius Quadratus aus Korinth, (in preparation).}. This kind of result is not achieved through an inquiry 
into a database but through a constructive consideration on the basis of personal 
experience. Such knowledge will remain also the basis of our discipline in the 
future. Databases complement our knowledge but they cannot replace it. This can't 
be their business.

Inscriptions are in the first place texts which contain certain evidence as such. 
But inscriptions do not only consist of text, although this may commonly seem to 
be the case in the literature. In German, one speaks of ``\textit{Inschriften 
errichten}'' and in Italy one can say: \textit{una iscrizione eretta}\footnote{It seems not necessary to give examples for this type of formulation; they are abundant.}. In the Prosopographia 
Imperii Romani,\textit{ }the expression \textit{titulum ponere }was long in use 
and applied to inscriptions of all kinds. With these expressions, it was not clear, 
what specific function was connected with an epigraphic text and the support on 
which the inscription appears. Nonetheless, the support already contains its own 
message and that was, on many occasions, the primary message for the Greek and 
Roman public\footnote{\label{fn:12}Cf. now W. Eck, Wie ehrt man Mitglieder der staatlich-städtischen Elite? Inschriftenbasen unerkannter Reiterdenkmäler, Kolloquium Kassel März 2015, hg. K. Ruffing – H. Schneider (in print)}. In order to understand what an inscription can today convey as evidence, 
it is fundamental to include the \textbf{monument }with which the inscription was 
connected. This can be an altar, an architrave of a temple, a mausoleum or, for 
instance, the base for a statue, which is explained by the inscription. In enquiring 
about the content of inscriptions, we muss decisively take into consideration the 
function that inscription and monument shared in the Graeco-Roman world.

However, when one looks for the function of an inscription in a database, the difficulties 
begin – if I were to speak about my own experiences. On many occasions, this 
function is not found, if one enquires about what is directly conveyed in the inscription, 
because this exact function is commonly not mentioned in the epigraphic text. This 
is not surprising. To mention the function was not particularly necessary because 
the ancient observer completely saw, on what an inscription was written. Given 
that today this context is mostly not preserved, we must infer it from the support 
whose form and particular details are connected with it. This necessity has meanwhile 
become almost common knowledge – in contrast with Mommsen's time\footnote{W. Eck, Mommsen e il metodo epigrafico, in: Concordia e la X Regio. Giornate di Studio in onore di Dario Bertolini, Atti del Convegno Portogruaro 22-23 ottobre 1994, Padova 1995, 107 ff.}. On the EAGLE 
website, this is specifically taken into account with the option \textit{advanced 
search, }in which it is possible to browse the database according to object type. 
With this option, one can find the categories, for example, of \textit{statue, 
statue base, statue base shaft or plinth}. Hence it is already possible to obtain 
a result with a high degree of differentiation, if one browses selectively. However, 
not all the different categories of statues are grasped with these search terms. 
Just to mention an example: \textit{bigae }or \textit{quadrigae, }which were erected 
to honour a person, can only be found if these terms would also appear in the inscription. 
But in the epigraphic text the object is directly mentioned only in a limited number 
of cases. The ancient observer saw the object which did not need any additional 
reference. For us, on the other hand, the specific types of statues are generally 
recognisable only if we include the support of the inscription in the analysis. 
The original statues are almost always lost or, at least, not connected with the 
inscription. Enquiring about the types of statues is not a child's play. For the 
most part, they rather say fundamental things for both the honoured and the honouring 
individuals.

One example from Volsinii (modern Bolsena) can illustrate this\footnote{AE 1980, 426.}. It is an honorific 
monument set up to commemorate a senator of the Hadrianic-Antonine age. The surviving 
fragmentary \textit{titulus honorarius }records his names: Pompeius Vopiscus C. 
Arruntius Catellius Celer Allius Sabinus, followed by his \textit{cursus }and, 
finally, the city of Carthage as dedicator. The text, however, does not specify 
in which figurative way the senator was presented to the audience of ancient Volsinii. 
This is also not attempted in the epigraphic database of Rome; there it is only 
said (as in the Epigraphischen Datenbank Heidelberg) that the text is to be read 
on a base\footnote{EDR077846 = HD005099 (M.G. Schmidt).}. While this is correct, the plain term ``base'' does not allow to recognise 
the most important element, namely with what type of honour had the city of Carthage 
honoured the senator coming from Volsinii. The particular quality of the honour 
is, nevertheless, revealed by the size and shape of the inscribed plaque. The plaque 
is 76.5 cm high and 1.90 m wide\footnote{P. Gros, MEFRA 92, 1980, 977 ff.}. The monument was so wide that a normal statue 
could not have stood on it; this would have appeared completely out of place. Given 
the width of the inscribed slab, it is to be inferred that a \textit{biga }or perhaps 
even a \textit{quadriga} was set up for the senator\textit{.} He should therefore 
have completed an action beyond his official duty on behalf of Carthage for which 
the city decided to recognise his activity in this lavish way and to send ambassadors 
to Volsinii to erect the monument.

A similar case can be inferred for Larcius Priscus, a legate of the \textit{legio 
III Augusta}\footnote{\emph{CIL VIII} 17891 = Dessau 1055.}. The city of Thamugadi honoured him in a similar way to that of Carthago 
for the senator in Volsinii. On the forum of the colony founded by Trajan, not 
a normal honorary statue not even an equestrian statue – as previously suspected\footnote{\label{note:18}J. Bergemann, Römische Reiterstatuen. Ehrendenkmäler im öffentlichen Bereich, Mainz 1990, 147 Nr. E 90:  ``equestrian statue''.} 
– was set up but rather a \textit{biga }on which the statue of Larcius Priscus 
was presented to the citizens of Timgad\footnote{The inscription is 188 cm wide, much more than necessary for an equestrian statue.}. But in EAGLE one finds the note:

Type of object: unbestimmt = undefined (= EDH)\footnote{EDH031159 (B. Gräf).}.

But one can see the type of the object on the photo, published by Zimmer in 1989\footnote{G. Wesch-Klein, Katalog der Inschriften, in: G. Zimmer, Locus datus decreto decurionum. Zur Statuenaufstellung zweier Forumsanlagen im römischen Africa, München 1989, 71 Nr. T 4.}. 
More examples of this kind could be given. In my opinion, an important consequence 
is therefore clear. If databases were to fulfil their function completely, then 
it would be not only useful but rather very necessary to record an indication of 
the concrete form of the honorific monument in such cases. Otherwise, a crucial 
part of the potential information would be lost.

Of course such a degree of precision is not always possible, but very likely when 
– as shown by the two previous examples – the form of the support provides 
evidence beyond the text of the inscription that can be searched in the database. 
This happens for example in many inscriptions which in antiquity were to be seen 
under \textit{statuae equestres}. Sometimes this form of honour is recorded in 
the text of the inscription itself. In the database Clauss, there are around 45 
records if one looks for \textit{statuae equestres }or \textit{statuam equestrem},\textit{ 
}in the database EAGLE around 27. That EAGLE shows fewer cases is naturally due 
to the fact that this database is still under construction. Either with 45 or 22 
inscriptions, this only records in any case a minority of the texts which were 
once to be read under such \textit{statuae equestres}. In the monograph on equestrian 
statues written by Johannes Bergemann in 1990, 128 inscriptions are collected, 
which were once connected with an equestrian statue according to the investigations 
of the author \footnote{See n. \ref{note:18}.}. Besides the statues directly referred to as such in the text, there 
are also those containing other hints which allow us to recognise that the honoured 
person was once presented on a horse.

Some examples should illustrate which equestrian statues cannot be recognised as 
such without these extra hints. In the colonia Aelia Capitolina, modern Jerusalem, 
an equestrian statue was erected for Antoninus Pius through the council of decurions, 
apparently straight after his accession in the year 138/9. In the text of the honorific 
inscription, the type of statue is not mentioned but it can be inferred from the 
size of the inscription\footnote{Corpus Inscriptionum Iudaeae/Palaestinae I 2, 718.}. Therefore, it must be registered as an equestrian statue 
in a database. The same applies to the honours bestowed in Lepcis Magna upon the 
governor of the late-antique province Tripolitana, Flavius Nepotianus. In the inscription 
itself, it is only spoken of a \textit{statua marmorea} that should memorialise 
the services of the magistrate\footnote{IRT 565.}. However, the base enables to recognise that a \textit{statua 
equestris }once stood there because this base is 170 cm deep\footnote{Bergemann (n. \ref{note:18}) no. 83. Cf. the inscription for the flamen provinciae Pompeius Cerealis Salvianus in Lepcis Magna: the basis is 70 cm high, but 160 cm deep (IRT 602).}.

\begin{figure}[!bp]
\centering
 \includegraphics[scale=0.5]{images/fig01ReiterstatueZeichnung.pdf}
\caption{Equestrian statue: drawing}
\label{fig:1}
\end{figure}

\begin{figure}[!bp]
\centering
\includegraphics[scale=0.5]{images/fig02ReiterstatueVonDerSeite.pdf}
\caption{Equestrian statue: view from the side}
\label{fig:2}
\end{figure}

>>>>

%&pdfLaTeX
% !TEX encoding = UTF-8 Unicode
\documentclass{article}
\usepackage{ifxetex}
\ifxetex
\usepackage{fontspec}
\setmainfont[Mapping=tex-text]{STIXGeneral}
\else
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\fi
\usepackage{textcomp}

\usepackage{amssymb}
\usepackage{fancyhdr}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}

\pagestyle{fancy}
\rhead{}
\rfoot{}
\chead{}
\cfoot{}
\lhead{\thepage{}}
\begin{document}

[2/13/2016 2:10 PM; 41021]

\begin{center}
\textbf{Tradition and progress.}

\textbf{The Roman World in the Digital Age - seen through Inscriptions. }
\end{center}

Epigraphy has caught up with the digital age. If this point needs to be proven, 
this very conference speaks for itself.  Its celebration in Rome does not happen 
by chance. Indeed, the first real attempt to create a collective compilation and 
presentation of all the Latin (and Greek) inscriptions in a single database or 
as a connection of the existing databases was made here in Rome in 1997 through 
Silvio Panciera during the 11\textsuperscript{th} International Congress of Greek 
and Latin Epigraphy. His fundamental initiative set in motion a process, which 
step by step will lead to a complete database. Very important steps in this direction 
have been made in the meantime and the development is progressing with particular 
emphasis at the Sapienza thanks to the collaboration between traditional epigraphists 
such as Silvia Orlandi and IT-people, who are affiliated to the Centro Linceo Beniamino 
Segre. These two and all the other partners in many European countries have to 
be thanked for these present steps and we wish them both courage and success in 
the following phase.

Whoever is dealing with inscriptions, at least since the 16\textsuperscript{th} 
century, has to deal with a continuously growing mass phenomenon. One of the reasons, 
which instigated Mommsen to establish the \textit{CIL}, was the amount of epigraphic 
texts, which were known at that time but had not been collected in one single place 
but were, instead, spread in countless manuscripts and publications. Even when 
one looks into a representative selection such as the \textit{Inscriptiones Latinae 
Selectae }by Hermann Dessau – completed in 1916 – there are more than 9500 
texts. Since the time of Mommsen and Dessau, however, the number of epigraphic 
documents has increased to a size previously unknown. On the website of EAGLE one 
can read of more than 1.5 million items, currently scattered across 25 EU countries 
as well as the east and south Mediterranean, which should be collected in the EAGLE 
database. In order to illustrate this growth in a more accessible way, let me refer 
to the following example. Since the conclusion of the first edition of \textit{CIL 
II}, namely with the supplement in 1892, the number of inscriptions from the Iberian 
Peninsula has tripled. For some types of inscriptions, the increase is even bigger. 
When Herbert Nesselhauf collected all the military diplomata in 1935, he could 
put together a total of 187 documents in \textit{CIL XVI. }Today, we know more 
than 1100 such texts, i.e. an increase of 7 times. Not in all regions of the early 
Roman Empire and for all types of inscriptions is the increase as dramatic as in 
the aforementioned examples. Even so, a doubling can be identified in many Roman 
places. Just to mention some examples: in Sarmizegetusa, the first Roman colony 
following the conquest of Dacia, around 330 inscriptions were collected in the 
different volumes of \textit{CIL III.} Since then, more than the same number of 
new texts has appeared, perhaps around 400. The growth is almost always dramatic 
when intensive excavations lead to a massive increase in the amount of inscriptions 
available, as somehow happens in the cases of Sagalassos or Perge in Pamphylia. 
Perge provided only 4 Latin inscriptions for the \textit{CIL III} and today there 
are at least 42. Even more dramatic is the growth in Caesarea Maritima. In \textit{CIL 
}there were only 3 Latin inscriptions known from the capital of the province of 
Judea, while now we have in the Corpus Inscriptionum Iudaeae/Palaestinae more than 
270 Latin documents.

Each of us is familiar with this massive increase in epigraphic monuments and we 
are reminded of this every year by the always more extensive volumes of \textit{AE 
}and \textit{SEG}. Even if it is impossible to provide an exact number of all the 
Latin inscriptions, one can still take as a starting point the fact that there 
are approximately 500.000 texts available for our work, including the so-called 
\textit{instrumentum domesticum}. There are more than 495.000 inscriptions in the 
Database Clauss-Slaby. Around 72.000 we can find in the Heidelberg Database and 
around the same number in the Epigraphic Database of Rome, from which the Project 
EAGLE emerged, not to mention other participants as, among others, \textit{Hispania 
Epigraphica }or the Bari Epigraphic Database. With regard to the Greek inscriptions, 
it is much more difficult to give an approximate number because their digital record 
is not as advanced as in the case of the Latin inscriptions. As for inscriptions 
in other languages – such as Punic, Hebrew-Aramaic, Demotic or Nabatean – let 
me exclude them from this discussion, even though they are just as part of the 
epigraphic tradition of the Greek and Roman antiquity as Latin and Greek inscriptions 
– we should not forget this very important fact. Our work as historians or philologists 
is based on all those texts. But, the bigger the number of documents providing 
information is, the more difficult it becomes to find all texts. Even more difficult 
is to collect selectively those texts related to the topic on which one wants to 
work. And yet, the complete collection is crucial for the scientific result.

@@

Is that different today because there are numerous databases available? At least 
with regard to the Latin inscriptions almost all of them are collected, although 
with very different individual information. With few clicks, we can find an answer 
for many questions in few seconds – sometimes one has to wait a bit longer, since 
the server does not work properly or some other problems exist. Indeed, as an already 
deceased colleague once told me long before the digital period in 1973: \textit{Pazienza} 
is the greatest virtue of the epigraphist. Today we are ready to forget this virtue 
thanks to our computers. At any rate, most of the relevant material available in 
databases appears quickly on the screen. It is not necessary to give examples. 
And yet, something which we all know: it is by no means guaranteed that we have 
found absolutely all the relevant material which exists; the reason for these gaps 
are inter alia, that not a small part of our epigraphic testimonies survives only 
as fragments and can be restored in different ways. For example, if we would try 
to find all the \textit{quaestores }of the province of Achaea in the imperial age, 
we would not find the following fragmentary inscription from Corinth, the capital 
of the province, because the text has read as follows:

\textit{Q(uinto) Vili[o --] / Titia[no] / Quadra[to --] / IIIIvir(o) vi[ar(um) 
cur(andarum), trib(uno)] / la}\textsuperscript{\textit{┌}}\textit{t}\textsuperscript{\textit{┐}}\textit{icl(avio) 
}\textsuperscript{\textit{┌}}\textit{le}\textsuperscript{\textit{┐}}\textit{g(ionis) 
[--, }\textit{\textbf{leg(ato) prov(inciae)] / Ach(aiae),}}\textit{ ob ius[titiam 
et fidem,] / quam circa [universam] / Achaia[m exhibuit,] / qu}\textsuperscript{\textit{┌}}\textit{a}\textsuperscript{\textit{┐}}\textit{m 
pro pr(aetore) [administravit.] / }\textsuperscript{\textit{┌}}\textit{M}\textsuperscript{\textit{┐}}\textit{(arcus) 
An}\textsuperscript{\textit{┌}}\textit{to}\textsuperscript{\textit{┐}}\textit{nius 
M(arci) [f(ilius)] / Nigrinus [--].}

Our search in a database would have recorded a proconsular legate of Achaea whose 
function in the inscription is nevertheless restored. The restoration of the position 
of provincial legate is not impossible, but it is also not very likely. The restoration 
of the office of quaestor:\textit{ [-- quaest(ori) prov(inciae)] / Ach(aiae) }instead 
of \textit{[-- leg(ato) prov(inciae)] / Ach(aiae) }fits in the lacuna much better. 
For a young senator who had completed an office of the \textit{vigintiviri }and 
then a tribunate in the legion, the \textit{quaestura }should actually follow in 
a normal career and not the position of a proconsular legate under a proconsul. 
Of course, we know that exceptionally young senators already before the \textit{quaestura 
}were acting as legates of a proconsul in his province. In total, there are only 
four of such cases. By contrast, much more senators state in their inscriptions 
that they have held the office of provincial \textit{quaestor} as\textit{ pro praetore}. 
The \textit{pro pr(aetore) }preserved in the inscription could therefore refer 
to the \textit{quaestura}. As the \textit{quaestura }normally follows the\textit{ 
}viginvirate and the legionary tribunate, this rule should automatically be observed 
in a fragmentary career. This kind of result is not achieved through an inquiry 
into a database but through a constructive consideration on the basis of personal 
experience. Such knowledge will remain also the basis of our discipline in the 
future. Databases complement our knowledge but they cannot replace it. This can't 
be their business.

Inscriptions are in the first place texts which contain certain evidence as such. 
But inscriptions do not only consist of text, although this may commonly seem to 
be the case in the literature. In German, one speaks of \textit{„Inschriften 
errichten``}and in Italy one can say: \textit{una iscrizione eretta}. In the Prosopographia 
Imperii Romani,\textit{ }the expression \textit{titulum ponere }was long in use 
and applied to inscriptions of all kinds. With these expressions, it was not clear, 
what specific function was connected with an epigraphic text and the support on 
which the inscription appears. Nonetheless, the support already contains its own 
message and that was, on many occasions, the primary message for the Greek and 
Roman public. In order to understand what an inscription can today convey as evidence, 
it is fundamental to include the \textbf{monument }with which the inscription was 
connected. This can be an altar, an architrave of a temple, a mausoleum or, for 
instance, the base for a statue, which is explained by the inscription. In enquiring 
about the content of inscriptions, we muss decisively take into consideration the 
function that inscription and monument shared in the Graeco-Roman world.

However, when one looks for the function of an inscription in a database, the difficulties 
begin – if I were to speak about my own experiences. On many occasions, this 
function is not found, if one enquires about what is directly conveyed in the inscription, 
because this exact function is commonly not mentioned in the epigraphic text. This 
is not surprising. To mention the function was not particularly necessary because 
the ancient observer completely saw, on what an inscription was written. Given 
that today this context is mostly not preserved, we must infer it from the support 
whose form and particular details are connected with it. This necessity has meanwhile 
become almost common knowledge – in contrast with Mommsen's time. On the EAGLE 
website, this is specifically taken into account with the option \textit{advanced 
search, }in which it is possible to browse the database according to object type. 
With this option, one can find the categories, for example, of \textit{statue, 
statue base, statue base shaft or plinth}. Hence it is already possible to obtain 
a result with a high degree of differentiation, if one browses selectively. However, 
not all the different categories of statues are grasped with these search terms. 
Just to mention an example: \textit{bigae }or \textit{quadrigae, }which were erected 
to honour a person, can only be found if these terms would also appear in the inscription. 
But in the epigraphic text the object is directly mentioned only in a limited number 
of cases. The ancient observer saw the object which did not need any additional 
reference. For us, on the other hand, the specific types of statues are generally 
recognisable only if we include the support of the inscription in the analysis. 
The original statues are almost always lost or, at least, not connected with the 
inscription. Enquiring about the types of statues is not a child's play. For the 
most part, they rather say fundamental things for both the honoured and the honouring 
individuals.

One example from Volsinii (modern Bolsena) can illustrate this. It is an honorific 
monument set up to commemorate a senator of the Hadrianic-Antonine age. The surviving 
fragmentary \textit{titulus honorarius }records his names: Pompeius Vopiscus C. 
Arruntius Catellius Celer Allius Sabinus, followed by his \textit{cursus }and, 
finally, the city of Carthage as dedicator. The text, however, does not specify 
in which figurative way the senator was presented to the audience of ancient Volsinii. 
This is also not attempted in the epigraphic database of Rome; there it is only 
said (as in the Epigraphischen Datenbank Heidelberg) that the text is to be read 
on a base. While this is correct, the plain term ``base'' does not allow to recognise 
the most important element, namely with what type of honour had the city of Carthage 
honoured the senator coming from Volsinii. The particular quality of the honour 
is, nevertheless, revealed by the size and shape of the inscribed plaque. The plaque 
is 76.5 cm high and 1.90 m wide. The monument was so wide that a normal statue 
could not have stood on it; this would have appeared completely out of place. Given 
the width of the inscribed slab, it is to be inferred that a \textit{biga }or perhaps 
even a \textit{quadriga} was set up for the senator\textit{.} He should therefore 
have completed an action beyond his official duty on behalf of Carthage for which 
the city decided to recognise his activity in this lavish way and to send ambassadors 
to Volsinii to erect the monument.

A similar case can be inferred for Larcius Priscus, a legate of the \textit{legio 
III Augusta.} The city of Thamugadi honoured him in a similar way to that of Carthago 
for the senator in Volsinii. On the forum of the colony founded by Trajan, not 
a normal honorary statue not even an equestrian statue – as previously suspected 
– was set up but rather a \textit{biga }on which the statue of Larcius Priscus 
was presented to the citizens of Timgad. But in EAGLE one finds the note:

Type of object: unbestimmt = undefined (= EDH).

But one can see the type of the object on the photo, published by Zimmer in 1989. 
More examples of this kind could be given. In my opinion, an important consequence 
is therefore clear. If databases were to fulfil their function completely, then 
it would be not only useful but rather very necessary to record an indication of 
the concrete form of the honorific monument in such cases. Otherwise, a crucial 
part of the potential information would be lost.

Of course such a degree of precision is not always possible, but very likely when 
– as shown by the two previous examples – the form of the support provides 
evidence beyond the text of the inscription that can be searched in the database. 
This happens for example in many inscriptions which in antiquity were to be seen 
under \textit{statuae equestres}. Sometimes this form of honour is recorded in 
the text of the inscription itself. In the database Clauss, there are around 45 
records if one looks for \textit{statuae equestres }or \textit{statuam equestrem},\textit{ 
}in the database EAGLE around 27. That EAGLE shows fewer cases is naturally due 
to the fact that this database is still under construction. Either with 45 or 22 
inscriptions, this only records in any case a minority of the texts which were 
once to be read under such \textit{statuae equestres}. In the monograph on equestrian 
statues written by Johannes Bergemann in 1990, 128 inscriptions are collected, 
which were once connected with an equestrian statue according to the investigations 
of the author. Besides the statues directly referred to as such in the text, there 
are also those containing other hints which allow us to recognise that the honoured 
person was once presented on a horse.

Some examples should illustrate which equestrian statues cannot be recognised as 
such without these extra hints. In the colonia Aelia Capitolina, modern Jerusalem, 
an equestrian statue was erected for Antoninus Pius through the council of decurions, 
apparently straight after his accession in the year 138/9. In the text of the honorific 
inscription, the type of statue is not mentioned but it can be inferred from the 
size of the inscription. Therefore, it must be registered as an equestrian statue 
in a database. The same applies to the honours bestowed in Lepcis Magna upon the 
governor of the late-antique province Tripolitana, Flavius Nepotianus. In the inscription 
itself, it is only spoken of a \textit{statua marmorea} that should memorialise 
the services of the magistrate. However, the base enables to recognise that a \textit{statua 
equestris }once stood there because this base is 170 cm deep.

Equestrian statues are, nevertheless, not uniform. They can also provide, in turn, 
additional information by their specific appearance. They were dedicated in very 
different forms; above all in very varied sizes. For example: Alfius Secundus, 
a \textit{flamen perpetuus }in Africa proconsularis, set up two or three equestrian 
statues of the emperor Septimius Severus. Even though these statues represent the 
emperor, they must have been small ones because the base is only 54 cm high and 
35 cm wide. These small equestrian statues fitted probably into the context in 
which they stood. On the other hand, gigantic equestrian statues were erected of 
all the emperors – and not only of them. An exact relation between the size of 
a monument and social rank did not exist. Various factors could be relevant in 
such cases. Nonetheless, the size can already tell us something about the status 
of the honorand and the intention of the dedicator.
